Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-04-08 writeups/Su_IGW_break.tex
try add revisions

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We approximate the equation of state as
\textcolor{Corr}{barotropic}, i.e.\ $P(\rho, T) = P(\rho)$. We study dynamics in
2D, so that fluid variables depend only on the Cartesian coordinates $x$ and
$z$. \textcolor{Corr}{While it is well known that waves break differently in two
versus three dimensions} \citep{klostermeyer,winters1994}, the dynamical effect
of the breaking process is likely to be similar in 2D \citep{barker_ogilvie}. We
approximate the gravitational field as uniform, pointing in the $(-\uv{z})$
direction. The plane-parallel approximation is justified since wave breaking
generally occurs near the stellar surface. The background density stratification
is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $-g\uv{z}$. For this isothermal background, hydrostatic
equilibrium implies \textcolor{Corr}{$\overline{P}(z) = \overline{\rho}(z) g
H$}. We assume there is initially no background flow, so $\bm{u} = \bm{u}'$.
Physically, this assumption corresponds to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}.
\textcolor{Corr}{These equations admit the canonical IGW solution}, given by
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        \textcolor{Corr}{The additional $\p{2H}^{-2}$ term in the denominator
        accounts for the growing amplitude of the IGW in the $z$ direction (the
        $z$ wavenumber is effectively $k_z - i / (2H)$).}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}}
        \approx 0$. In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. \textcolor{Corr}{Any incident horizontal
momentum flux absorbed by the fluid, denoted $F_a(t)$, must manifest as
additional horizontal momentum of the shear flow}.
\textcolor{Corr}{Additionally, as the mean flow $\overline{U}$ cannot grow
efficiently above $\overline{U}_c$ (due to the breakdown of the linear
solution), we assume $\overline{U}$ saturates at $\overline{U}_c$, which holds
to good accuracy (see Fig.~\ref{fig:nl_fluxes}). Instead, the critical layer
must propagate downward in response to the incident momentum flux.} The
horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z /
u_{z})$, where the minimum is taken over every grid point in the domain, and
$\Delta x \equiv L_x / N_x$ and $\Delta z \equiv L_z / N_z$ are the grid
spacings in the $x$ and $z$ directions respectively.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. \textcolor{Corr}{Most importantly, horizontal momentum is still
conserved between $z_B$ and $z_T$}. We choose the transition width $\Delta z =
0.25H$ and damping time $\tau = 1 / (15N)$. This prescription is similar to
\citet{lecoanet_damp} and has the advantage of being smooth, important for
spectral methods. Further details of our implementation of the fluid equations
in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
solution $\bm{u}_{al}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to analytical solution $\bm{u}_{al}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{al}$, then
$A_i(t) = 1$. \textcolor{Corr}{The factor of $\bar{\rho}$ inside the integrands
in Eq.~\eqref{eq:ahat_def} corrects for the $\propto e^{z/2H}$ growth of
$\bm{u}_{al}$; without it, $A_i(t)$ would be dominated by the contribution near
$z_t$.}

@@

The analytical theory (Section~\ref{ss:lin_analysis}) also predicts that the
horizontal momentum flux $F(z, t)$ is independent of $z$ between the forcing
zone where the wave is generated and the damping zone where it is dissipated.
The expected horizontal momentum flux carried by the excited IGW in the linear
theory can be computed by simply evaluating Eq.~\eqref{eq:F_def} for
$\bm{u}_{al}$ and is a constant:
\begin{equation}
    F_{al} \equiv \ev{\rho u_{al, x} u_{al, z}}_x.\label{eq:F_al}
\end{equation}
Denoting the momentum flux measured in the simulation by $F(z, t)$
(use Eq.~\eqref{eq:F_def} with velocities taken from the simulation), we expect
$F(z, t) = F_{al}$ between $z_0$ and $z_T$. Fig.~\ref{fig:lin_fluxes} shows
agreement with this prediction.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_fluxes.png}
    \caption{$F/F_{al}$ plotted at select times $t$ (in units of
    $N^{-1}$). As the initial transient dies out, $F / F_{al} \approx 1$ to a
    good approximation above the forcing zone $z > z_0 = 2H$ and below the
    damping zone $z \lesssim z_T = 9.5H$. The horizontal momentum flux excited
    in the forcing zone is transported without loss to the top of the domain,
    where it is dissipated by the damping zone (see Section~\ref{ss:damping})
    without reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

    \item At the end of the simulation (bottom right panel), the
        \textcolor{Corr}{critical layer has advanced downwards, but otherwise}
        the flow shows very few significant qualitative differences from the
        previous snapshot. This suggests that the latter phase of the simulation
        has reached a steady state.

        \textcolor{Corr}{Notably, the horizontal banded structure of $u_x$ in
        the synchronized fluid is mostly unchanged (also visible in the top
        panel of Fig.~\ref{fig:nl_fluxes}), suggesting that momentum
        redistribution and mixing within the synchronized fluid are negligible.}
\end{enumerate}

@@

In Fig.~\ref{fig:nl_fluxes}, we plot the mean horizontal flow velocity
$\overline{U}$ (Eq.~\eqref{eq:mean_flow}) and the dimensionless momentum flux $F
/ F_{al}$ (Eqs.~\eqref{eq:F_def} and~\eqref{eq:F_al}) as a function of $z$ at
the times depicted in Fig.~\ref{fig:snapshots}. At each time, $\overline{U}$ is
close to zero below the critical layer, but then sharply increases to
$\overline{U}_c$ at the critical layer (i.e.\ the flow is ``spun-up''). Above
the critical layer, $\overline{U}$ varies slightly due to momentum transport
within the spun-up layer. This agrees with the expectation discussed in
Section~\ref{ss:crit_layer}.

Similarly, $F \lesssim F_{al}$ below the critical layer, and then decreases to
about zero above the critical layer. However, two notable deviations from the
discussion in Section~\ref{ss:crit_layer} can be observed: (i) the incident flux
on the critical layer fluctuates somewhat temporally, and (ii) there is a small
negative flux just above the critical layer at later times. These are addressed
in subsequent sections.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_fluxes.png}
    \caption{The mean horizontal flow velocity $\overline{U}(z, t)$
    (Eq.~\eqref{eq:mean_flow}) and the dimensionless momentum flux $F(z, t) /
    F_{al}$ (Eqs.~\eqref{eq:F_def} and~\eqref{eq:F_al}) in our fiducial
    simulation plotted at the same times as in Fig.~\ref{fig:snapshots}. The two
    distinct zones of mean flow are separated by a critical layer. The critical
    layer propagates toward lower $z$ due to momentum transport
    ($\pdil{F}{z}$).}\label{fig:nl_fluxes}
\end{figure}

@@

This result suggests that the critical layer width is regulated by the
competition between steepening induced by IGW breaking and broadening due to
shear instability. This width does not vary significantly with resolution in our
resolved simulations (see Fig.~\ref{fig:agg}). As such, $\mathrm{Ri} \sim 0.5$
can be used to calculate the critical layer width in stars, where $N^2$ and
$\overline{U}_c$ (corresponding to the tidal frequency) are known.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The red and solid green lines denote respectively the minimum
    and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean and
    spread in width of the critical layer over $x$. Note that $\mathrm{Ri} \sim
    \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at the
    critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

@@

After measuring $z_c$ (see Section~\ref{ss:cl_prop}) and $F(z)$
(Eq.~\eqref{eq:F_def}) at each time step, we determine each of $F_i$, $F_a$,
$F_r$, $F_s$ as follows:
\begin{align}
    F_i(t) &= F_{al}A_i^2(t),\\
    F_r(t) &= F_i(t) - \frac{1}{H}
        \int\limits_{z_c - \Delta z - H}^{z_c - \Delta z}F(z, t)\;\mathrm{d}z
        % \ev{F(z, t)}_{z \in [z_c - \Delta z - H, z_c - \Delta z]}
        ,\label{eq:fr_def}\\
    F_s(t) &=
        \frac{1}{\Delta z}\int\limits_{z_c}^{z_c + \Delta z}
            \min(F(z, t), 0)\;\mathrm{d}z,
        % \ev{\z{F(z, t): F(z, t) < 0}}_{z \in [z_c, z_c + \Delta z]},
        \label{eq:fs_def}\\
    F_a(t) &= F_i - F_r - F_s.\label{eq:fa_def}
\end{align}
Fig.~\ref{fig:nl_f_amps2} depicts the four components of this flux
decomposition.
% Here, $\ev{\dots}_{z \in [z_a, z_b]}$ denotes a vertical average over the
% interval $[z_a, z_b]$.
Below the critical layer, we average over an interval of length $H$, also the
vertical wavelength. The offset $\Delta z$ is necessary to make the measurement
of the incident flux unaffected by the turbulence within the critical layer
itself. The width of the critical layer is limited by $\mathrm{Ri} \lesssim 1$
(see Section~\ref{ss:khi}), which bounds its vertical extent $\sim
\frac{1}{\abs{k_{z}}}$. We empirically found an offset of $\Delta z =
\frac{3}{\abs{k_z}}$ was necessary to be sufficiently far from strong
fluctuations near the critical layer.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_amps2.png}
    \caption{Momentum flux decomposition calculated from the simulation. Plotted
    are the four components of the horizontal momentum flux budget over time
    (see Eq.~\eqref{eq:f_budget}), in units of the analytical estimate for the
    incident wave flux $F_{al}$ (Eq.~\eqref{eq:S_lin}): $F_i$, the flux incident
    on the critical layer (green); $F_a$, the flux absorbed by the critical
    layer (blue); $F_r$, the flux reflected at the critical layer (red); and
    $F_s$, the flux inside the synchronized layer
    (black).}\label{fig:nl_f_amps2}
\end{figure}

@@

With a careful determination of $F_a$, we can make predictions for the
propagation of $z_c(t)$ and compare to the measured propagation in the
simulation. In principle, $z_c$ is the location where the incident flux
significantly attenuates. In the simulation, shear turbulence causes $F$ to have
significant spatial and temporal fluctuations that translate to large temporal
fluctuations in $z_c(t)$. To minimize these spurious fluctuations, we measure
the location of the critical layer using a spatial average of where flux
deposition occurs:
\begin{align}
    z_{c, \min}(t) &\equiv \min_z \z{z: F(z, t) > 0.3F_{al}},\\
    z_{c, \max}(t) &\equiv \max_z \z{z: F(z, t) < 0.3F_{al}},\\
    z_c(t) &\equiv \frac{z_{c, \min}(t) + z_{c, \max}(t)}{2}.\label{eq:zc_def}
\end{align}
Measuring $z_c$ in other ways does not significantly change the results of the
analysis.

>>>>

We consider a incompressible, isothermally stratified fluid representing a
stellar envelope or atmosphere. We study dynamics in 2D, so that fluid variables
depend only on the Cartesian coordinates $x$ and $z$. \textcolor{Corr}{While it
is well known that waves break differently in 2D versus 3D} \citep{klostermeyer,
winters1994}, the dynamical effect of the breaking process is likely to be
similar in 2D \citep{barker_ogilvie}. We approximate the gravitational field as
uniform, pointing in the $(-\uv{z})$ direction. The plane-parallel approximation
is justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible fluid in a uniform gravitational field
are\footnote{\textcolor{Corr}{These equations arise by enforcing fluid
incompressibility ($\mdil{\rho}{t} = 0$ or the adiabatic index $\gamma \to
\infty$) in the Euler equations. They also can be derived from the commonly used
``sound-proof'' \emph{anelastic equations} \citep[see e.g.][]{ogura1962scale,
sutherland1} in the limit where (i) the density stratification is weak ($H \gg
L$ for perturbation wavelength $L$) and (ii) the fluid is incompressible (note
the anelastic equations are defined for arbitrary $\gamma$). Under condition
(i), the anelastic approximation reduces to the Boussinesq approximation
\citep{klein2009asymptotics}. Our equations permit a weak but nonzero density
stratification and can be interpreted as ``locally Boussinesq''. Condition (ii)
is motivated by WD fluid properties and, more generally, simplicity, as the
dynamical effect of wave breaking is likely to be similar for different sets of
fluid equations.}}
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $(-g\uv{z})$.

For this isothermal background, hydrostatic equilibrium implies
\textcolor{Corr}{$\overline{P}(z) = \overline{\rho}(z) g H$}. We assume there is
initially no background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}.
\textcolor{Corr}{The resulting equations admit the canonical IGW solution}
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        \textcolor{Corr}{The additional $\p{2H}^{-2}$ term in the denominator
        accounts for the growing amplitude of the IGW in the $z$ direction (as
        the $z$ wavenumber is effectively $k_z - i / (2H)$).} We note
        $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}} \approx 0$.
        In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. \textcolor{Corr}{Any incident horizontal
momentum flux absorbed by the fluid, denoted $F_a(t)$, must manifest as
additional horizontal momentum of the shear flow}.
\textcolor{Corr}{Additionally, as the mean flow $\overline{U}$ cannot grow
efficiently above $\overline{U}_c$ (due to the breakdown of the linear
solution), we assume $\overline{U}$ saturates at $\overline{U}_c$, which holds
to good accuracy (see Fig.~\ref{fig:nl_fluxes}). In this case, the critical
layer must propagate downward in response to the incident momentum flux.} The
horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from \textcolor{Corr}{the minimum of $0.1 / N$
and the advective Courant-Friedrichs-Lewy (CFL) time. The CFL time is given by}
$\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{z})$, where the minimum is
taken over every grid point in the domain, and $\Delta x \equiv L_x / N_x$ and
$\Delta z \equiv L_z / N_z$ are the grid spacings in the $x$ and $z$ directions
respectively.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. \textcolor{Corr}{Most importantly, horizontal momentum remains
conserved between $z_B$ and $z_T$, and outgoing boundary conditions are imposed
at $z_B, z_T$}. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
analytical solution $\bm{u}_{an}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to $\bm{u}_{an}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{an}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{an}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{an}$, then
$A_i(t) = 1$. \textcolor{Corr}{The factor of $\bar{\rho}$ inside the integrands
in Eq.~\eqref{eq:ahat_def} corrects for the $\propto e^{z/2H}$ growth of
$\bm{u}_{an}$; without it, $A_i(t)$ would be dominated by the contribution near
$z_t$.}

@@

The analytical theory (Section~\ref{ss:lin_analysis}) also predicts that the
horizontal momentum flux $F(z, t)$ is independent of $z$ between the forcing
zone where the wave is generated and the damping zone where it is dissipated.
The expected horizontal momentum flux carried by the excited IGW in the linear
theory can be computed by simply evaluating Eq.~\eqref{eq:F_def} for
$\bm{u}_{an}$ and is a constant:
\begin{equation}
    F_{an} \equiv \ev{\rho u_{an, x} u_{an, z}}_x.\label{eq:F_al}
\end{equation}
Denoting the momentum flux measured in the simulation by $F(z, t)$
(use Eq.~\eqref{eq:F_def} with velocities taken from the simulation), we expect
$F(z, t) = F_{an}$ between $z_0$ and $z_T$. Fig.~\ref{fig:lin_fluxes} shows
agreement with this prediction.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/lin_fluxes.png}
    \caption{$F/F_{an}$ plotted at select times $t$ (in units of
    $N^{-1}$). As the initial transient dies out, $F / F_{an} \approx 1$ to a
    good approximation above the forcing zone $z > z_0 = 2H$ and below the
    damping zone $z \lesssim z_T = 9.5H$. The horizontal momentum flux excited
    in the forcing zone is transported without loss to the top of the domain,
    where it is dissipated by the damping zone (see Section~\ref{ss:damping})
    without reflection.}\label{fig:lin_fluxes}
\end{figure}

@@

    \item At the end of the simulation (bottom right panel), the
        \textcolor{Corr}{critical layer has advanced downwards, but otherwise}
        the flow shows very few significant qualitative differences from the
        previous snapshot. This suggests that the latter phase of the simulation
        has reached a steady state. \textcolor{Corr}{Notably, the horizontal
        banded structure of $u_x$ in the upper, synchronized fluid does not
        continue to evolve (also visible in the top panel of
        Fig.~\ref{fig:nl_fluxes}), suggesting that momentum redistribution and
        mixing within the synchronized fluid are negligible.}
\end{enumerate}

@@

In Fig.~\ref{fig:nl_fluxes}, we plot the mean horizontal flow velocity
$\overline{U}$ (Eq.~\eqref{eq:mean_flow}) and the dimensionless momentum flux $F
/ F_{an}$ (Eqs.~\eqref{eq:F_def} and~\eqref{eq:F_al}) as a function of $z$ at
the times depicted in Fig.~\ref{fig:snapshots}. At each time, $\overline{U}$ is
close to zero below the critical layer, but then sharply increases to
$\overline{U}_c$ at the critical layer (i.e.\ the flow is ``spun-up''). Above
the critical layer, $\overline{U}$ varies slightly due to momentum transport
within the spun-up layer. This agrees with the expectation discussed in
Section~\ref{ss:crit_layer}.

Similarly, $F \lesssim F_{an}$ below the critical layer, and then decreases to
about zero above the critical layer. However, two notable deviations from the
discussion in Section~\ref{ss:crit_layer} can be observed: (i) the incident flux
on the critical layer fluctuates somewhat temporally, and (ii) there is a small
negative flux just above the critical layer at later times. These are addressed
in subsequent sections.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_fluxes.png}
    \caption{The mean horizontal flow velocity $\overline{U}(z, t)$
    (Eq.~\eqref{eq:mean_flow}) and the dimensionless momentum flux $F(z, t) /
    F_{an}$ (Eqs.~\eqref{eq:F_def} and~\eqref{eq:F_al}) in our fiducial
    simulation plotted at the same times as in Fig.~\ref{fig:snapshots}. The two
    distinct zones of mean flow are separated by a critical layer. The critical
    layer propagates toward lower $z$ due to momentum transport
    ($\pdil{F}{z}$).}\label{fig:nl_fluxes}
\end{figure}

@@

This result suggests that the critical layer width is regulated by the
competition between steepening induced by IGW breaking and broadening due to
shear instability. This width does not vary significantly with resolution in our
resolved simulations (see Fig.~\ref{fig:agg}). As such, $\mathrm{Ri} \sim 0.5$
can be used to calculate the critical layer width in stars, where $N^2$ and
$\overline{U}_c$ (corresponding to the tidal frequency) are known.
\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_ri.png}
    \caption{Local Richardson number (Eq.~\eqref{eq:ri_med_def}) of the flow at
    the critical layer over time (in units of $N^{-1}$) in our fiducial
    simulation. The solid red and dotted black lines denote respectively the
    minimum and median of $\mathrm{Ri}_x(x, t)$. These numbers measure the mean
    and spread in width of the critical layer over $x$. Note that $\mathrm{Ri}
    \sim \frac{1}{4}$ corresponds to the KHI, so this plot suggests the shear at
    the critical layer does not steepen past the onset of the
    KHI.}\label{fig:nl_f_ri}
\end{figure}

@@

After measuring $z_c$ (see Section~\ref{ss:cl_prop}) and $F(z)$
(Eq.~\eqref{eq:F_def}) at each time step, we determine each of $F_i$, $F_a$,
$F_r$, $F_s$ as follows:
\begin{align}
    F_i(t) &= F_{an}A_i^2(t),\\
    F_r(t) &= F_i(t) - \frac{1}{H}
        \int\limits_{z_c - \Delta z - H}^{z_c - \Delta z}F(z, t)\;\mathrm{d}z
        % \ev{F(z, t)}_{z \in [z_c - \Delta z - H, z_c - \Delta z]}
        ,\label{eq:fr_def}\\
    F_s(t) &=
        \frac{1}{\Delta z}\int\limits_{z_c}^{z_c + \Delta z}
            \min(F(z, t), 0)\;\mathrm{d}z,
        % \ev{\z{F(z, t): F(z, t) < 0}}_{z \in [z_c, z_c + \Delta z]},
        \label{eq:fs_def}\\
    F_a(t) &= F_i - F_r - F_s.\label{eq:fa_def}
\end{align}
Fig.~\ref{fig:nl_f_amps2} depicts the four components of this flux
decomposition.
% Here, $\ev{\dots}_{z \in [z_a, z_b]}$ denotes a vertical average over the
% interval $[z_a, z_b]$.
Below the critical layer, we average over an interval of length $H$, also the
vertical wavelength. The offset $\Delta z$ is necessary to make the measurement
of the incident flux unaffected by the turbulence within the critical layer
itself. The width of the critical layer is limited by $\mathrm{Ri} \lesssim 1$
(see Section~\ref{ss:khi}), which bounds its vertical extent $\sim
\frac{1}{\abs{k_{z}}}$. We empirically found an offset of $\Delta z =
\frac{3}{\abs{k_z}}$ was necessary to be sufficiently far from strong
fluctuations near the critical layer.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{plots/nl_f_amps2.png}
    \caption{Momentum flux decomposition calculated from the simulation. Plotted
    are the four components of the horizontal momentum flux budget over time
    (see Eq.~\eqref{eq:f_budget}), in units of the analytical estimate for the
    incident wave flux $F_{an}$ (Eq.~\eqref{eq:S_lin}): $F_i$, the flux incident
    on the critical layer; $F_a$, the flux absorbed by the critical layer;
    $F_r$, the flux reflected at the critical layer; and $F_s$, the flux inside
    the synchronized layer.}\label{fig:nl_f_amps2}
\end{figure}

@@

With a careful determination of $F_a$, we can make predictions for the
propagation of $z_c(t)$ and compare to the measured propagation in the
simulation. In principle, $z_c$ is the location where the incident flux
significantly attenuates. In the simulation, shear turbulence causes $F$ to have
significant spatial and temporal fluctuations that translate to large temporal
fluctuations in $z_c(t)$. To minimize these spurious fluctuations, we measure
the location of the critical layer using a spatial average of where flux
deposition occurs:
\begin{align}
    z_{c, \min}(t) &\equiv \min_z \z{z: F(z, t) > 0.3F_{an}},\\
    z_{c, \max}(t) &\equiv \max_z \z{z: F(z, t) < 0.3F_{an}},\\
    z_c(t) &\equiv \frac{z_{c, \min}(t) + z_{c, \max}(t)}{2}.\label{eq:zc_def}
\end{align}
Measuring $z_c$ in other ways does not significantly change the results of the
analysis.
